---
title: "Exploratory Data Analysis: Diabetes Health Indicators"
author: "Taylor Burke"
format:
  html:
    toc: true
    toc-depth: 3
execute:
  echo: true
  warning: false
  message: false
---
```{r setup}
#| label: setup
#| include: true

# Load required packages
library(tidyverse)   # data wrangling, plots
library(janitor)     # clean variable names
library(skimr)       # data summaries

# Set a reproducible seed
set.seed(123)
```

## Introduction

In this analysis, I explore the Diabetes Health Indicators dataset
(diabetes_binary_health_indicators_BRFSS2015.csv), which is based on the
Behavioral Risk Factor Surveillance System (BRFSS) survey.

- Response variable: Diabetes_binary
  - 0 = No diabetes
  - 1 = Diabetes
- Goal of EDA:
  - Understand the distribution of key health indicators.
  - Explore how these indicators relate to diabetes status.
  - Prepare for building predictive models in the Modeling document.

I focus on the following predictors (a subset of the full data):
- HighBP – high blood pressure
- HighChol – high cholesterol
- BMI – body mass index (numeric)
- Smoker – has smoked at least 100 cigarettes
- PhysActivity – physical activity in past 30 days
- GenHlth – general health (1 = excellent to 5 = poor)
- MentHlth – days of poor mental health
- PhysHlth – days of poor physical health
- DiffWalk – serious difficulty walking
- Sex – biological sex
- Age – age category
- Income – income category

These variables will form the basis of my later predictive modeling.

## Data Import and Cleaning


```{r}
#| label: import-data
#| message: true

# Read data from folder
raw_diabetes <- read_csv(
"diabetes_binary_health_indicators_BRFSS2015.csv"
)

# Clean column names to be snake_case
diabetes <- raw_diabetes %>%
clean_names()

# Inspect the structure
glimpse(diabetes)
```
### Converting Variables to Factors

Many columns are coded as 0/1 or numeric categories, but conceptually are categorical.
Here I convert them to factors with meaningful labels where appropriate.

```{r}
#| label: factor-conversion

diabetes <- diabetes %>%
mutate(
# Response as factor
diabetes_binary = factor(diabetes_binary,
levels = c(0, 1),
labels = c("NoDiabetes", "Diabetes")),
high_bp   = factor(high_bp,   levels = c(0, 1),
                   labels = c("NoHighBP", "HighBP")),
high_chol = factor(high_chol, levels = c(0, 1),
                   labels = c("NoHighChol", "HighChol")),
smoker    = factor(smoker,    levels = c(0, 1),
                   labels = c("NonSmoker", "Smoker")),
phys_activity = factor(phys_activity, levels = c(0, 1),
                       labels = c("NoPA", "PhysActive")),
diff_walk = factor(diff_walk, levels = c(0, 1),
                   labels = c("NoDiffWalk", "DiffWalk")),
sex       = factor(sex,       levels = c(0, 1),
                   labels = c("Female", "Male")),

# General health coded 1–5 (1 = excellent, 5 = poor)
gen_hlth = factor(gen_hlth,
                  levels = 1:5,
                  labels = c("Excellent", "VeryGood", "Good",
                             "Fair", "Poor")),

# Age categories and income categories as ordered factors
age = factor(age, ordered = TRUE),
income = factor(income, ordered = TRUE)
)

# Quick look after conversions
diabetes %>%
select(diabetes_binary, high_bp, high_chol, smoker,
phys_activity, diff_walk, sex, gen_hlth, age, income) %>%
summary()
```
### Missingness

Before any analysis, I check for missing values.

```{r}
#| label: missingness

# Count missing values per column

missing_summary <- diabetes %>%
summarise(across(everything(), ~ sum(is.na(.)))) %>%
pivot_longer(everything(),
names_to = "variable",
values_to = "n_missing") %>%
arrange(desc(n_missing))

missing_summary
```

## Overall Outcome Balance

First, I check how many respondents are labeled with diabetes vs. no diabetes.

```{r}
#| label: outcome-balance

diabetes %>%
count(diabetes_binary) %>%
mutate(prop = n / sum(n))
```

### Bar Plot of Outcome
```{r}
#| label: outcome-plot
#| fig.cap: "Distribution of Diabetes Outcomes"

diabetes %>%
ggplot(aes(x = diabetes_binary)) +
geom_bar() +
labs(
x = "Diabetes Status",
y = "Count",
title = "Counts of Diabetes vs. No Diabetes"
)
```

This plot shows class imbalance (typically many more NoDiabetes than Diabetes cases),
which I must keep in mind when choosing modeling strategies and evaluation metrics such as log-loss.